---
title: "Mananjary cohort malaria epi analysis"
author: "Benny Rice"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set up

## Header info

- Benjamin Rice on behalf of the Princeton-CRS Research Team
- Processing malaria rapid diagnostic test data and associated data
- Cleaning data and visualizing data collected 2021-2022
- Aim 1: Visualize temporal trends in malaria infection
- Aim 2: Organize data to apply a HMM to better model time course of infection among sampled individuals
- Aim 3: Associate malaria infection outcome data with relevant covariates for plotting

## Loading necessary packages
```{r i1, include=FALSE}
library(tidyverse)
library(readr)
library(gridExtra)
library(reshape2)
library(patchwork)
library(lubridate)
```

# Playing with S6: Tidying Data

## Reading in data
```{r 2.01}
#Playing with S6
df1i <- readxl::read_excel("/Users/blrice/Documents/R GIT REPOS/2022_MNJ_TAZO/data_rdt/S6_DATA_ENTRY_20221126t.xlsx")
head(df1i)
#Dropping blank rows
df1i <- df1i %>% filter(!(is.na(unique_ind_id)))
str(df1i)
```

## Cleaning up columns

- First, need to rename columns to make easier to use with pivot_longer
- Using only "_" to separate time point from variable
```{r 2.02}
names(df1i) <- c("unique_ind_id", "site_code", "hh_id", "CODE", "dob_yr", "dob_mo", "dob_day", "sex", 
                 "rdt_T8", "visit_T8", "yr.visit_T8", "mo.visit_T8", "day.visit_T8", "obs_T8", 
                 "rdt_T7", "visit_T7", "yr.visit_T7", "mo.visit_T7", "day.visit_T7", "obs_T7", 
                 "rdt_T6", "visit_T6", "yr.visit_T6", "mo.visit_T6", "day.visit_T6", "obs_T6", 
                 "rdt_T5", "visit_T5", "yr.visit_T5", "mo.visit_T5", "day.visit_T5", "obs_T5", 
                 "rdt_T4", "visit_T4", "yr.visit_T4", "mo.visit_T4", "day.visit_T4", "obs_T4", 
                 "rdt_T3", "visit_T3", "yr.visit_T3", "mo.visit_T3", "day.visit_T3", "obs_T3", 
                 "rdt_T2", "visit_T2", "yr.visit_T2", "mo.visit_T2", "day.visit_T2", "obs_T2", 
                 "rdt_T1", "visit_T1", "yr.visit_T1", "mo.visit_T1", "day.visit_T1", "obs_T1")
```


- Formatting date columns into a sample.date column
```{r 2.03, warning=FALSE}
#Combining year, month, and date into a formatted sample.date column
df1 <- df1i %>% 
  #Converting "NA"s to NAs
  #To avoid doing for all columns one at a time, using across() to apply to a set of columns
  #Using "." to simply pass each column from before into the na_if() function 
  mutate(across(dob_yr:obs_T1, ~ na_if(., "NA"))) %>%
  #Adding a combined date column for each time_point using ymd()
  #Probably is a more elegant way to do this
  #Note because individuals not sampled at a time point have NAs, will be some dates that don't parse
  mutate(sample.date_T8 = ymd(paste0(yr.visit_T8, "-", mo.visit_T8, "-", day.visit_T8)),
         sample.date_T7 = ymd(paste0(yr.visit_T7, "-", mo.visit_T7, "-", day.visit_T7)),
         sample.date_T6 = ymd(paste0(yr.visit_T6, "-", mo.visit_T6, "-", day.visit_T6)),
         sample.date_T5 = ymd(paste0(yr.visit_T5, "-", mo.visit_T5, "-", day.visit_T5)),
         sample.date_T4 = ymd(paste0(yr.visit_T4, "-", mo.visit_T4, "-", day.visit_T4)),
         sample.date_T3 = ymd(paste0(yr.visit_T3, "-", mo.visit_T3, "-", day.visit_T3)),
         sample.date_T2 = ymd(paste0(yr.visit_T2, "-", mo.visit_T2, "-", day.visit_T2)),
         sample.date_T1 = ymd(paste0(yr.visit_T1, "-", mo.visit_T1, "-", day.visit_T1)))
#Dropping the unnecessary date columns, CODE columns
df1 <- df1 %>% select("unique_ind_id", "site_code", "hh_id", "dob_yr", "dob_mo", "dob_day", "sex", 
                      "rdt_T8", "sample.date_T8", "obs_T8", 
                      "rdt_T7", "sample.date_T7", "obs_T7", 
                      "rdt_T6", "sample.date_T6", "obs_T6", 
                      "rdt_T5", "sample.date_T5", "obs_T5", 
                      "rdt_T4", "sample.date_T4", "obs_T4", 
                      "rdt_T3", "sample.date_T3", "obs_T3", 
                      "rdt_T2", "sample.date_T2", "obs_T2", 
                      "rdt_T1", "sample.date_T1", "obs_T1")
```

- Fixing DOBs
- For now:
  - If month unknown, use June
  - If date unknown, use 15th
```{r 2.04}
#Fixing dobs
#Converting to numeric
#If month unknown, use June as a guess for now
#If date unknown, use 15th as a guess for now
df2 <- df1 %>% 
  mutate(dob_yr  = as.numeric(dob_yr)) %>%
  mutate(dob_mo  = as.numeric(dob_mo)) %>%
  mutate(dob_mo  = replace_na(dob_mo, 6)) %>%
  mutate(dob_day = as.numeric(dob_day)) %>%
  mutate(dob_day = replace_na(dob_day, 15)) %>%
  #Combining into a dob column, date format
  mutate(dob = ymd(paste0(dob_yr, "-", dob_mo, "-", dob_day))) %>%
  #Selecting columns
  select("unique_ind_id", "site_code", "hh_id", "dob", "sex", 
         "rdt_T8", "sample.date_T8", "obs_T8", 
         "rdt_T7", "sample.date_T7", "obs_T7", 
         "rdt_T6", "sample.date_T6", "obs_T6", 
         "rdt_T5", "sample.date_T5", "obs_T5", 
         "rdt_T4", "sample.date_T4", "obs_T4", 
         "rdt_T3", "sample.date_T3", "obs_T3", 
         "rdt_T2", "sample.date_T2", "obs_T2", 
         "rdt_T1", "sample.date_T1", "obs_T1")
```

## Tidying data: Pivoting longer

- Pivoting longer
```{r 2.05}
#Pivoting longer
df3 <- df2 %>%
  #Undoing date format to allow combining
  mutate(across(rdt_T8:obs_T1, ~ as.character(.))) %>%
  pivot_longer(
    #Specify the columns to tuck under
    #NOT The columns to stay at the left
    cols = !(unique_ind_id:sex),
    #Splitting the variable (e.g., rdt result) from the time point (e.g., T8)
    #Using the spacer "_" that separates variable name from time_point
    names_to = c("variable", "time_point"),
    names_sep = "_",
    #sending values to a column named value
    values_to = "value"
  )
head(df3)
```

- Too long: Pivoting wider to get columns for sample.date and observations

```{r 2.06}
#Pivoting wider to get columns for sample.date and observations
df4 <- df3 %>%
  pivot_wider(names_from = variable, values_from = value) %>%
  #converting sample.date back to date format
  mutate(sample.date = ymd(sample.date))
head(df4)
```

## Recoding RDT results

### Recoding as

    - inv (NA);
    - neg (0);
    - pan (1);
    - panpf (2);
    - pf (3)
    
```{r 2.07}
#Recoding RDT result
table(df4$rdt)
df5 <- df4 %>% mutate(rdt = case_when(
  rdt == "inv"   ~ NA_real_,
  rdt == "neg"   ~ 0,
  rdt == "pf"    ~ 1,
  rdt == "pan"   ~ 1,
  rdt == "panpf" ~ 1))
table(df5$rdt)
```

# Playing with S6: First plots

- X-axis: **time_point** or **real dates**; 
- Y-axis: unique_ind_id faceted by household; 
- Dots colored by infection status:
  - Absent/un-sampled: leave blank
  - Negative: grey
  - Positive: orange

```{r 3.01}
p3_01a.cols <- c("0" = "#8A877B", "1" = "#EB6429")
p3_01a <- df5 %>% 
  #Drop rows with no data
  filter(!is.na(rdt)) %>% filter(!is.na(sample.date)) %>%
  #Show X households at a time
  filter(hh_id %in% unique(df5$hh_id)[1:10]) %>%
  ggplot(aes(x = time_point, y = unique_ind_id, color = as.factor(rdt))) +
  geom_point() +
  scale_colour_manual(values = p3_01a.cols) +
  facet_grid(rows = vars(hh_id), scales = "free_y", space = "free") +
  xlab("Time") + ylab("Individuals") +
  labs(color="Infection status") +
  theme_bw() +
  theme(strip.background = element_rect(fill = "white"),
        axis.text.y = element_blank())

p3_01b <- df5 %>% 
  #Drop rows with no data
  filter(!is.na(rdt)) %>% filter(!is.na(sample.date)) %>%
  #Show X households at a time
  filter(hh_id %in% unique(df5$hh_id)[1:10]) %>%
  ggplot(aes(x = sample.date, y = unique_ind_id, color = as.factor(rdt))) +
  geom_point() +
  scale_colour_manual(values = p3_01a.cols) +
  facet_grid(rows = vars(hh_id), scales = "free_y", space = "free") +
  xlab("Time") + ylab("Individuals") +
  labs(color="Infection status") +
  theme_bw() +
  theme(strip.background = element_rect(fill = "white"),
        axis.text.y = element_blank())
```

## Plotting 10 HHs as an example {.tabset}

- In gray: negative RDT
- In orange: positive RDT
- No data: Absent, declined, or not yet born

### By Time Point

**Figure 3.01a**

- T1 Jul-Oct 2021 - T8 Oct-Nov 2022

```{r 3.01.1, fig.height=8}
p3_01a
```

### By Date

**Figure 3.01b**

- Site S6: First 10 households shown as an example

```{r 3.01.2, fig.height=8}
p3_01b
```


## Plotting all 50 HHs {.tabset}

```{r 3.02}
#creating plot objects for all
p_01_10 <- df5 %>% 
  #Drop rows with no data
  filter(!is.na(rdt)) %>% filter(!is.na(sample.date)) %>%
  #Show X households at a time
  filter(hh_id %in% unique(df5$hh_id)[1:10]) %>%
  ggplot(aes(x = sample.date, y = unique_ind_id, color = as.factor(rdt))) +
  geom_point() +
  scale_colour_manual(values = p3_01a.cols) +
  facet_grid(rows = vars(hh_id), scales = "free_y", space = "free") +
  xlab("Time") + ylab("Individuals") +
  labs(color="Infection status") +
  theme_bw() +
  theme(strip.background = element_rect(fill = "white"),
        axis.text.y = element_blank())

p_11_20 <- df5 %>% 
  #Drop rows with no data
  filter(!is.na(rdt)) %>% filter(!is.na(sample.date)) %>%
  #Show X households at a time
  filter(hh_id %in% unique(df5$hh_id)[11:20]) %>%
  ggplot(aes(x = sample.date, y = unique_ind_id, color = as.factor(rdt))) +
  geom_point() +
  scale_colour_manual(values = p3_01a.cols) +
  facet_grid(rows = vars(hh_id), scales = "free_y", space = "free") +
  xlab("Time") + ylab("Individuals") +
  labs(color="Infection status") +
  theme_bw() +
  theme(strip.background = element_rect(fill = "white"),
        axis.text.y = element_blank())

p_21_30 <- df5 %>% 
  #Drop rows with no data
  filter(!is.na(rdt)) %>% filter(!is.na(sample.date)) %>%
  #Show X households at a time
  filter(hh_id %in% unique(df5$hh_id)[21:30]) %>%
  ggplot(aes(x = sample.date, y = unique_ind_id, color = as.factor(rdt))) +
  geom_point() +
  scale_colour_manual(values = p3_01a.cols) +
  facet_grid(rows = vars(hh_id), scales = "free_y", space = "free") +
  xlab("Time") + ylab("Individuals") +
  labs(color="Infection status") +
  theme_bw() +
  theme(strip.background = element_rect(fill = "white"),
        axis.text.y = element_blank())

p_31_40 <- df5 %>% 
  #Drop rows with no data
  filter(!is.na(rdt)) %>% filter(!is.na(sample.date)) %>%
  #Show X households at a time
  filter(hh_id %in% unique(df5$hh_id)[31:40]) %>%
  ggplot(aes(x = sample.date, y = unique_ind_id, color = as.factor(rdt))) +
  geom_point() +
  scale_colour_manual(values = p3_01a.cols) +
  facet_grid(rows = vars(hh_id), scales = "free_y", space = "free") +
  xlab("Time") + ylab("Individuals") +
  labs(color="Infection status") +
  theme_bw() +
  theme(strip.background = element_rect(fill = "white"),
        axis.text.y = element_blank())

p_41_50 <- df5 %>% 
  #Drop rows with no data
  filter(!is.na(rdt)) %>% filter(!is.na(sample.date)) %>%
  #Show X households at a time
  filter(hh_id %in% unique(df5$hh_id)[41:50]) %>%
  ggplot(aes(x = sample.date, y = unique_ind_id, color = as.factor(rdt))) +
  geom_point() +
  scale_colour_manual(values = p3_01a.cols) +
  facet_grid(rows = vars(hh_id), scales = "free_y", space = "free") +
  xlab("Time") + ylab("Individuals") +
  labs(color="Infection status") +
  theme_bw() +
  theme(strip.background = element_rect(fill = "white"),
        axis.text.y = element_blank())

p_51_58 <- df5 %>% 
  #Drop rows with no data
  filter(!is.na(rdt)) %>% filter(!is.na(sample.date)) %>%
  #Show X households at a time
  filter(hh_id %in% unique(df5$hh_id)[51:58]) %>%
  ggplot(aes(x = sample.date, y = unique_ind_id, color = as.factor(rdt))) +
  geom_point() +
  scale_colour_manual(values = p3_01a.cols) +
  facet_grid(rows = vars(hh_id), scales = "free_y", space = "free") +
  xlab("Time") + ylab("Individuals") +
  labs(color="Infection status") +
  theme_bw() +
  theme(strip.background = element_rect(fill = "white"),
        axis.text.y = element_blank())
```

### HHs 01_10
```{r 3.02.1, fig.height=8}
p_01_10
```

### HHs 11_20
```{r 3.02.2, fig.height=8}
p_11_20
```

### HHs 21_30
```{r 3.02.3, fig.height=8}
p_21_30
```

### HHs 31_40
```{r 3.02.4, fig.height=8}
p_31_40
```

### HHs 41_50
```{r 3.02.5, fig.height=8}
p_41_50
```

### HHs 51_58
```{r 3.02.6, fig.height=7}
p_51_58
```




